---
title: "How to fit the AccelerAge framework in R: an illustration"
author: "Marije Sluiskes"
output: rmarkdown::github_document
---

This document contains a step-by-step tutorial on how to fit the AccelerAge framework. It has two sections: in the first section I simulate a synthetic lifetable and dataset, in the second section I provide a step-by-step illustration of the AccelerAge framework. Those primarily interested in the AccelerAge model fitting procedure can skip the first section and move immediately to the second one.

```{r, echo=F, warning=F, message=F}
library(survival)
library(scam)
library(ggplot2)
library(dplyr)
library(eha)
library(survminer)
library(gridExtra)
source("src/gompertz_draw.R")

set.seed(36) # run entire script at once for reproducible results
```

# Simulate the data needed for the illustration

I simulate the data according to the procedure described in the paper, assuming an Accelerated Failure Time (AFT) model with a Gompertz baseline.

The technicalities of the data generation procedure are described in the paper. In this illustration I use mean residual life $m(t)$ as the measure of remaining lifetime. Given that the subject has survived until time $t$, and conditional on their marker values $X$, mean residual life is defined as: $m(t|X) = E[T - t| T > t, X] = \int_t^{\infty} \frac{S(u|X)}{S(t|X)}du$. 

Under the AFT assumption, $S(t|X)$ is given by $S_0[t \times \exp(\beta^TX)]$.The factor $\exp(\beta^TX)$ can hence directly be interpreted as an individual aging rate: if it is greater than 1, a subject experiences accelerated aging. If it is smaller than 1, decelerated aging. 

```{r}
################################################################
#### Parameters
################################################################

a <- exp(-9)
b <- 0.085
sigma <- 1/b # canonical parametrization
tau <- a/b # canonical parametrization

followup = 20
M <- 2 # number of predictors (all standard normally distributed)
betas <- c(0.05, 0.05) # vector of beta's (should be of length M)

N_pop <- 1e5 # population size (from which lifetable is determined)
n_obs <- 5e3 # number of observations 

################################################################
#### Create lifetable for population
################################################################

X <- matrix( rnorm(N_pop*M,mean=0,sd=1), N_pop, M) # matrix of predictors 

t <- vector(length = N_pop)                                                     
for (i in 1:N_pop){
  t[i] <- rgompertz_aft(1, sigma = sigma, tau = tau, linpred = sum(betas * X[i,])) # vector of ages-of-death
}

lifetable_pop <- as.data.frame(cbind(X, t))
lifetable_pop <- lifetable_pop[order(lifetable_pop$t),]

mrl <- vector(length = N_pop)
for (j in 1:N_pop){
  mrl[j] <- mean(lifetable_pop$t[j:N_pop]) - lifetable_pop$t[j]
}

lifetable_pop$mrl <- mrl

# smoothen 
fitsmooth <- scam(mrl ~ s(t, bs = "mpd"), data = lifetable_pop)
xx <- seq(0,max(lifetable_pop$t), by = 0.1)
lt <- as.data.frame(cbind(t = xx, mrl =  predict(fitsmooth, data.frame(t=xx))))

################################################################
#### Create data set 
################################################################

n_gen <- 2 * n_obs # twice as many to ensure I generate enough, because for some T < C => not observed

X <- as.data.frame(matrix( rnorm(n_gen*M,mean=0,sd=1), n_gen, M))
colnames(X) <- c("x1", "x2")
age_start <- runif(n_gen, 20, 80)
linpred <- rowSums(sweep(X, 2, betas, "*"))

# Get age of death
age_death <- vector(length = n_gen)
for (i in 1:n_gen){
  age_death[i] <- rgompertz_aft(1, sigma = sigma, linpred = linpred[i], tau = tau)
}

# Remove observations that are left-truncated
indx_obs <- which(age_start < age_death)[1:n_obs]  # [1:n_obs[k]] to obtain the intended sample size 
df_sim <- as.data.frame(cbind(X, age_death, age_start, linpred))[indx_obs,]

# Get mean residual life
for (i in 1:nrow(df_sim)){
  mrl_uncon <- integrate(gomp_baseline_surv, lower = (df_sim$age_start[i] * exp(df_sim$linpred[i])), 
                         upper=Inf, a = a, b = b)$value
  s_cond <-  gomp_baseline_surv(df_sim$age_start[i] * exp(df_sim$linpred[i]), a = a, b = b) 
  df_sim$mrl[i] <- (mrl_uncon / s_cond) * exp(-df_sim$linpred[i])
}

# Get biological age (via population lifetable)
for (i in 1:nrow(df_sim)){
  df_sim$b_age[i] <- lt$t[ which.min(abs(lt$mrl - df_sim$mrl[i])) ]
}

# Add censoring 
df_sim$yrs_rem <- df_sim$age_death - df_sim$age_start
wh <- which(df_sim$yrs_rem > followup) # censored
df_sim$status <- 1
df_sim$status[wh] <- 0
df_sim$follow_up_time <- df_sim$yrs_rem
df_sim$follow_up_time[wh] <- followup
df_sim$age_end <- df_sim$age_start + df_sim$follow_up_time
```

## Plots of the simulated data
```{r, echo=F}
mycol = "darkorchid1"
# lifetable
lt %>% 
  ggplot(aes(x = t, y = mrl)) +
  geom_line(col = mycol, size = 1) +
  labs(title = "Lifetable", x = "chronological age", y = "mean residual life") +
  coord_cartesian(xlim = c(0, 100)) +
  theme_minimal()

# biological age 
df_sim %>% 
  ggplot(aes(x = age_start, y = b_age)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, col = mycol, size = 1) + 
  labs(title = "Chronological and biological age", 
       x = "chronological age", y = "biological age") +
  theme_minimal()

# ages of death
df_sim %>% 
  ggplot(aes(x = age_death, y = ..density..)) +
  geom_histogram(binwidth = 1) + 
  labs(title = "Ages of death (conditional on being alive at age of inclusion)", 
       x = "chronological age", y = "density") +
  theme_minimal()

# Kaplan-Meier plot 
fitlt <- survfit(Surv(age_start, age_end, status) ~ 1, data = df_sim)
fitlt %>% 
  ggsurvplot(xlim = c(20,100),
           palette = mycol,
           ggtheme = theme_minimal()) +
  labs(title = "Kaplan-Meier plot of simulated dataset (df_sim)")


```

# Fitting the AccelerAge framework

In this section the AccelerAge framework is fitted on the data simulated above. To fit the AccelerAge-framework, you need:

* A dataset `df_sim` with for each subject: markers of interest (here: `x1` and `x2`), chronological age at entry (here: `age_start`) and the chronological age at which the subject dies or is censored (here: `age_end`). 
* A lifetable of your population of interest (here: `lt`). This lifetable should contain the mean residual life value within the population for different values of chronological age. (If you are interested in some other function of residual life, e.g. median residual life, it should contain these values.)

If you are fitting a parametric AFT model, as I am in this illustration, you will also need to make an assumption about the underlying distribution of the event times. In this case I will fit an AFT Gompertz model, which is in correspondence with the true data-generating mechanism. 

For simplicity I do not use a test data set, but if you have enough data available (or, even better, an external validation data set) it would be advisable to fit your model on a different data set than the one you are evaluating it on. 

### 1. Fit the AFT model and obtain the parameter estimates

I use the `aftreg` function from the [`eha` package](https://cran.r-project.org/web/packages/eha/eha.pdf). This function can fit an AFT Gompertz model and is able to deal with delayed entry data (left-truncated data).

```{r}
fit_aft <- eha::aftreg(formula = Surv(age_start, age_end, status) ~ x1 + x2, 
                          data = df_sim, dist = "gompertz") 

# get estimated parameters
sigma_est <- exp(fit_aft$coefficients["log(scale)"]) # canonical parametrization
tau_est <- exp(fit_aft$coefficients["log(shape)"]) # canonical parametrization
a_est <- tau_est / sigma_est
b_est <- 1 / sigma_est
```

If instead of an AFT Gompertz model, you want to fit a parametric AFT model with a different distribution, you can change the `dist` parameter (e.g. to `weibull`, `loglogistic` or `lognormal`).

### 2. Obtain the linear predictors

The linear predictors can be obtained by multiplying your predictor variables, `x1` and `x2` here, with the estimated parameter coefficients. 

```{r}
betas_est_aft <- fit_aft$coefficients[1:M]
linpred_aft <- rowSums(sweep(df_sim[,1:M], 2, betas_est_aft, "*"))
```


### 3. Predict mean residual life

Your regression will have resulted in estimated Gompertz parameters ($\hat{a}$ and $\hat{b}$, or in the [canonical parametrization](https://cran.r-project.org/web/packages/eha/vignettes/gompertz.html), $\hat{\sigma}$ and $\hat{\tau}$). By integrating over the Gompertz survival curve defined by these parameters, taking into account the linear predictors, it is possible to obtain a predicted mean residual life value.

```{r}
for (i in 1:nrow(df_sim)){
  mrl_uncon <- integrate(gomp_baseline_surv, lower = (df_sim$age_start[i] * exp(linpred_aft[i])), 
                         upper=Inf, a = a_est, b = b_est)$value
  s_cond <-  gomp_baseline_surv(df_sim$age_start[i] * exp(linpred_aft[i]), a = a_est, b = b_est) 
  df_sim$pred_mrl[i] <- (mrl_uncon / s_cond) * exp(-linpred_aft[i])
}
```

In this simulated scenario, I know the true mean residual life values. So I can compare my predicted values with the true values. This would of course not be possible in a real data setting.

```{r, echo=F}
df_sim %>% 
  ggplot(aes(x = mrl, y = pred_mrl)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, col = mycol, size = 1) +
  labs(title = "Comparison true and predicted mean residual life values",
       x = "true mean residual life", y = "predicted mean residual life")
```

They agree very closely, which is in line with what I want to see: after all, my dataset `df_sim` was simulated under an AFT Gompertz scenario. 

### 4. Obtain the biological age predictions

In this final step the lifetable is needed to move from mean residual life to biological age. Given a mean residual life prediction, I select the chronological age in the lifetable whose corresponding mean residual life value is closest to the prediction.

```{r}
for (i in 1:length(df_sim$pred_mrl)){
  df_sim$pred_b_age[i] <- lt$t[which.min(abs(lt$mrl - df_sim$pred_mrl[i]))]
}
```

Once again, since I know the underlying true biological age in this simulated scenario, I can check whether my predicted biological age values correspond with the true ones. When using real data this is not an option. In that case you can look at measures such as calibration or discrimination, as elaborated upon in the paper. 

```{r}
df_sim %>% 
  ggplot(aes(x = b_age, y = pred_b_age)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, col = mycol, size = 1) +
  labs(title = "Comparison true and predicted biological ages",
       x = "true biological age", y = "predicted biological age")
```




